Cross-Loop Optimization of Arithmetic Intensity for Finite Element Local Assembly

### 2014-09-10 TACO-paper-v2/coffee.tex
Incorporate gpu paragraph. Several other fixes

\begin{abstract}
We study and systematically evaluate a class of composable code
transformations that improve arithmetic intensity in local assembly
operations, which represent a significant fraction of the execution
time in finite element methods. Their performance optimization is indeed a
challenging issue. Even though affine loop nests are generally present,
the short trip counts and the complexity of mathematical expressions, which
vary among different problems, make it hard to determine a single or unique
sequence of successful transformations. Our investigation has resulted in the
implementation of a compiler for local assembly kernels, fully
integrated with a framework for developing finite element methods.
The compiler manipulates abstract syntax trees generated from a domain-specific
language by introducing domain-aware optimizations and, eventually,
produces C code including vector SIMD intrinsics. Experiments using a range of finite-element
problems of increasing complexity show that significant performance improvement
is achieved. The generality of the approach and the applicability of the proposed
code transformations to other domains is also discussed.

@@

@@

Achieving high performance is non-trivial. The complexity of the
mathematical expressions, often characterized by a large number of
operations on constants and small matrices, makes it hard to determine a
single or specific sequence of transformations that is successfully
applicable to all problems. Loop trip counts are typically small and can
vary significantly, which further exacerbates the issue. A
compiler-based approach is, therefore, the only reasonable option to
obtain close-to-peak performance in a wide range of different local
assembly kernels. Optimizations like padding, generalized loop-invariant
code motion, vector-register tiling, and expression splitting, as well
as their composition, are essential, but their support in
state-of-the-art polyhedral and vendor compilers, if any, is very
limited. BLAS routines could theoretically be employed, although a
fairly complicated control and data flow analysis would be required to
automate identification and extraction of matrix-matrix multiplies. In
addition, as detailed in Section~\ref{sec:perf-eval-blas}, the small
dimension of the matrices involved and the potential loss in data
locality can limit or eliminate the performance gain of this approach.

@@

\begin{figure}
\begin{alltt}
\scriptsize
\textbf{Input:} element matrix (2D array, initialized to 0), coordinates (array), coefficients (array, e.g. velocity)
\textbf{Output:} element matrix (2D array)
- Compute Jacobian from coordinates
- Define basis functions
- Compute element matrix in an affine loop nest
\end{alltt}
\caption{Structure of a local assembly kernel}
\label{code:general-structure}
\end{figure}

@@

% Register allocation and register locality
\subsection{Model-driven Vector-register Tiling}
\label{sec:op-vect}
One notable problem of assembly kernels concerns register allocation
and register locality. The critical situation occurs when loop trip
counts and the variables accessed are such that the vector-register
pressure is high. Since the kernel's working set fits the L1 cache, it
is particularly important to optimize register management. Standard
optimizations, such as loop interchange, unroll, and unroll-and-jam,
can be employed to deal with this problem. In COFFEE, these
optimizations are supported either by means of explicit code
transformations (interchange, unroll-and-jam) or indirectly by
delegation to the compiler through standard pragmas (unroll). Tiling
at the level of vector registers is an additional feature of
COFFEE. Based on the observation that the evaluation of the element
matrix can be reduced to a summation of outer products along the
\texttt{j} and \texttt{k} dimensions, a model-driven vector-register
tiling strategy can be implemented. If we consider the code snippet in
Listing~\ref{code:helmholtz-licm} and we ignore the presence of the
operation \texttt{det*W3[i]}, the computation of the element matrix is
abstractly expressible as
\begin{equation}
\label{outer-product}
A_{jk} = \sum_{\substack{
  x \in B' \subseteq B \\
  y \in B'' \subseteq B}}
x_j\cdot y_k ~~~~~~ j,k = 0,...,2
\end{equation}
where $B$ is the set of all basis functions (or temporary variables,
e.g., \texttt{LI$\_$0}) accessed in the kernel, whereas $B'$ and $B''$
are generic problem-dependent subsets. Regardless of the specific
input problem, by abstracting from the presence of all variables
independent of both \texttt{j} and \texttt{k}, the element matrix
computation is always reducible to this form.
Figure~\ref{fig:vect-by-vect} illustrates how we can evaluate 16 entries
($j,k=0,...,3$) of the element matrix using just 2 vector registers,
which represent a 4$\times$4 tile, assuming $\vert B' \vert = \vert B''
\vert = 1$. Values in a register are shuffled each time a product is
performed. Standard compiler auto-vectorization for both GNU and Intel
compilers, instead, executes 4 broadcast operations (i.e., ``splat'' of
a value over all of the register locations) along the outer dimension to
perform the calculation. In addition to incurring a larger number of
cache accesses, it needs to keep between $f=1$ and $f=3$ extra registers
to perform the same 16 evaluations when unroll-and-jam is used, with $f$
being the unroll-and-jam factor.

@@

% FIXME: ``mesh used to discretize''
%\rule[0.2ex]{16cm}{0.5pt}
\begin{figure}[b]
\begin{center}
\includegraphics[scale=0.75]{Pictures/coffee-pipeline.pdf}
\caption{High-level view of Firedrake. COFFEE is at the core,
  receiving ASTs from a modified version of the FEniCS Form Compiler
  and producing optimized C code kernels.}
\label{fig:coffee-pipeline}
\end{center}
\end{figure}

@@

\subsubsection{Choice and Properties of the Benchmarks}
\label{sec:real-examples}
Our code transformations were evaluated in three real-world problems
based on the following PDEs:
\begin{enumerate}
\item Helmholtz
\item Advection-Diffusion
\item Burgers
\end{enumerate}
The three chosen benchmarks are \emph{real-life kernels} and comprise
the core differential operators in some of the most frequently
encountered finite element problems in scientific computing. This is
of crucial importance because distinct problems, possibly arising  in
completely different fields, may employ (subsets of) the same differential
operators of our benchmarks, which implies similarities
and redundant patterns in the generated code. Consequently, the proposed
code transformations have a domain of applicability that goes far beyond
that of the three analyzed equations.

@@

\begin{table*}[t]
\tbl{Performance improvement due to generalized loop-invariant code
  motion, for different element shapes (triangle, tetrahedron, prism)
  and polynomial orders ($p \in [1, 4]$), over the original
  non-optimized code, for the Helmholtz, Diffusion and Burgers
  problems.}{ \scriptsize \licmresultsnorms }
\label{table:perf-results-licm}
\end{table*}
Table~\ref{table:perf-results-licm} illustrates the performance
improvement obtained when \emph{licm} is applied. In general, the
speedups are notable. The main reasons were anticipated in
Section~\ref{sec:code-transf-licm}: in the original code, 1)
sub-expressions invariant to outer loops are not automatically
hoisted, while 2) sub-expressions invariant to the innermost loop are
hoisted, but their execution is not auto-vectorized. These
observations come from inspection of assembly code generated by the
compiler.

@@

\begin{table*}[t]
\tbl{Performance improvement due to generalized loop-invariant code
  motion, data alignment, and padding, for different element shapes
  (triangle, tetrahedron, prism) and polynomial orders ($p \in [1,
    4]$), over the original non-optimized code, for the Helmholtz,
  Diffusion and Burgers problems.}{ \scriptsize \licmapresultsnorms }
\label{table:perf-results-licmap}
\end{table*}

@@

\begin{figure}[t]
	\centering
	\subfloat[Sandy Bridge]{	\includegraphics[scale=0.52]{Pictures/helmholtz-normalized-opvect.pdf}}
	\subfloat[Xeon Phi]{	\includegraphics[scale=0.52]{Pictures/helmholtz-normalized-opvect-phi.pdf}}
	\caption{Performance improvement over \emph{licm-ap} obtained
          by \emph{op-vect} in the Helmholtz kernel. Bars suffixed
          with ``CM'' indicate that the cost model was used to
          transform the kernel.}
	\label{fig:opvect-helmholtz-speedup}
\end{figure}

\begin{figure}
	\centering
	\subfloat[Sandy Bridge]{	\includegraphics[scale=0.52]{Pictures/diffusion-normalized-opvect.pdf}}
	\subfloat[Xeon Phi]{	\includegraphics[scale=0.52]{Pictures/diffusion-normalized-opvect-phi.pdf}}
	\caption{Performance improvement over \emph{licm-ap} obtained
          by \emph{op-vect} in the Diffusion kernel. Bars suffixed
          with ``CM'' indicate that the cost model was used to
          transform the kernel.}
	\label{fig:opvect-diffusion-speedup}
\end{figure}

@@

\begin{figure}
	\centering
	\subfloat[Sandy Bridge]{	\includegraphics[scale=0.52]{Pictures/burgers-normalized-split.pdf}}
	\subfloat[Xeon Phi]{	\includegraphics[scale=0.52]{Pictures/burgers-normalized-split-phi.pdf}}
	\caption{Performance improvement over \emph{licm-ap} obtained
          by \emph{split} in the Burgers kernel. Bars suffixed with
          ``CM'' indicate that the cost model was used to transform
          the kernel.}
	\label{fig:split-burgers-speedup}
\end{figure}

@@

\begin{table}[h]
\tbl{Performance comparison between FEniCS and COFFEE on the Sandy Bridge.}{
\begin{tabulary}{1.0\columnwidth}{C|C|C}
\hline
Problem & Max slow down & Max speed up \\\hline
Helmholtz & - & 4.14$\times$ \\
Diffusion & - & 4.28$\times$ \\
Burgers & 2.24$\times$ & 1.61$\times$ \\\hline
\end{tabulary}
}
\label{table:comparison-to-FFC-opt}
\end{table}

@@

\begin{figure}
	\centering
	\subfloat[]{\includegraphics[scale=0.52]{Pictures/elasticity-p1.pdf}}
	\subfloat[]{\includegraphics[scale=0.52]{Pictures/elasticity-p2.pdf}}
	\caption{Performance improvement over non-optimized code for the static linear elasticity equation. The tetrahedral mesh is composed of 196608 elements. Experiments were executed on a single core of a Sandy Bridge architecture.}
	\label{fig:full-app-res}
\end{figure}

@@

\section{Generality of the Approach and Applicability to Other Domains}
\label{sec:generality}
We have demonstrated that our cross-loop optimizations for arithmetic intensity are effective in the context of automated code generation for finite element local assembly. In this section, we discuss about their applicability in other computational domains and, more in general, their integrability within a general-purpose compiler.

COFFEE was developed as a separate, self-contained software module, with
clear input/output interfaces, rather then incorporating it within
PyOP2. This choice was motivated by two critical aspects that
characterize the generality of our research.

@@

\item The various sub-expressions in $\beta$ only depend on (i.e. iterate along) a subset of the enclosing loops. In addition, some of these sub-expressions might reduce to the same values as iterating along certain iteration spaces. This code structure motivated the generalized loop-invariant code motion technique. The intuition is that whenever sub-expressions invariant with respect to different sets of affine loops can be identified, the question of whether, where and how to hoist them, while minimizing redundant computation, arises. Pre-computation of invariant terms also increases memory requirements due to the need for temporary arrays, so it is possible that for certain architectures the transformation could actually cause slow downs (e.g. whenever the available per-core memory is small).

@@

  transformations. For instance, knowing loop bounds allows both generating

@@

  Section~\ref{sec:op-vect}) and discovering redundant computation

@@

  general-purpose compiler is far from straightforward, especially if stencils are present, we believe that this could be more easily achieved in specific domains. The intuition is to map the memory access pattern on vector registers, and then exploit in-register shuffling to minimize the traffic between memory and processor. By demonstrating the effectiveness of multi-loop vectorization in a real scenario, our research represents an incentive for studying this technique in a broader and systematic way.

>>>>

\begin{abstract}
We study and systematically evaluate a class of composable code
transformations that improve arithmetic intensity in local assembly
operations, which represent a significant fraction of the execution
time in finite element methods. Their performance optimization is indeed a
challenging issue. Even though affine loop nests are generally present,
the short trip counts and the complexity of mathematical expressions, which
vary among different problems, make it hard to determine a single or unique
sequence of successful transformations. Our investigation has resulted in the
implementation of a compiler for local assembly kernels, fully
integrated with a framework for developing finite element methods.
The compiler manipulates abstract syntax trees generated from a domain-specific
language by introducing domain-aware optimizations and, eventually,
produces C code including vector SIMD intrinsics. Experiments using a range of real-world finite-element
problems of increasing complexity show that significant performance improvement
is achieved. The generality of the approach and the applicability of the proposed
code transformations to other domains is also discussed.

@@

We aim to maximize our impact on the platforms that are realistically used
for finite element applications, so we target conventional CPU architectures. 
The key limiting factor to the execution on GPUs is the stringent memory requirements. 
Only relatively small meshes can fit a GPU memory, and distributed execution is minimally
supported by available frameworks (if any). There has been some research on
adapting local assembly to GPUs (which we mention in Section~\ref{sec:related-work}),
although it differs from ours in several ways, including: (i) not relying on
automated code generation from a domain-specific language (explained next),
(ii) testing only very low order methods, (iii) not optimizing for
cross-loop arithmetic intensity (the goal is rather effective multi-thread
parallelization). In addition, our code transformations would drastically impact the 
GPU parallelization strategy, for example by increasing a thread's working set. 
For all these reasons, a study on extending our work to GPU architectures is 
beyond the scope of this paper. In Section~\ref{sec:generality}, however, we provide 
some intuitions about this research direction.

@@

Achieving high performance on CPUs is non-trivial. The complexity of the
mathematical expressions, often characterized by a large number of
operations on constants and small matrices, makes it hard to determine a
single or specific sequence of transformations that is successfully
applicable to different problems (equations). Loop trip counts are typically small and can
vary significantly, which further exacerbates the issue. A
compiler-based approach is, therefore, the only reasonable option to
obtain close-to-peak performance in a wide range of different local
assembly kernels. Optimizations like padding, generalized loop-invariant
code motion, vector-register tiling, and expression splitting, as well
as their composition, are essential, but their support in
state-of-the-art polyhedral and vendor compilers, if any, is very
limited. BLAS routines could theoretically be employed, although a
fairly complicated control and data flow analysis would be required to
automate identification and extraction of matrix-matrix multiplies. In
addition, as detailed in Section~\ref{sec:perf-eval-blas}, the small
dimension of the matrices involved and the potential loss in data
locality can limit or eliminate the performance gain of this approach.

@@

\begin{figure}[b]
\begin{alltt}
\scriptsize
\textbf{Input:} element matrix (2D array, initialized to 0), coordinates (array), coefficients (array, e.g. velocity)
\textbf{Output:} element matrix (2D array)
- Compute Jacobian from coordinates
- Define basis functions
- Compute element matrix in an affine loop nest
\end{alltt}
\caption{Structure of a local assembly kernel}
\label{code:general-structure}
\end{figure}

@@

% Register allocation and register locality
\subsection{Model-driven Vector-register Tiling}
\label{sec:op-vect}
One notable problem of assembly kernels concerns register allocation
and register locality. The critical situation occurs when loop trip
counts and the variables accessed are such that the vector-register
pressure is high. Since the kernel's working set fits the L1 cache, it
is particularly important to optimize register management. Standard
optimizations, such as loop interchange, unroll, and unroll-and-jam,
can be employed to deal with this problem. In COFFEE, these
optimizations are supported either by means of explicit code
transformations (interchange, unroll-and-jam) or indirectly by
delegation to the compiler through standard pragmas (unroll). Tiling
at the level of vector registers is an additional feature of
COFFEE. Based on the observation that the evaluation of the element
matrix can be reduced to a summation of outer products along the
\texttt{j} and \texttt{k} dimensions, a model-driven vector-register
tiling strategy can be implemented. If we consider the code snippet in
Listing~\ref{code:helmholtz-licm} and we ignore the presence of the
operation \texttt{det*W3[i]}, the computation of the element matrix is
abstractly expressible as
\begin{equation}
\scriptsize
\label{outer-product}
A_{jk} = \sum_{\substack{
  x \in B' \subseteq B \\
  y \in B'' \subseteq B}}
x_j\cdot y_k ~~~~~~ j,k = 0,...,2
\end{equation}
where $B$ is the set of all basis functions (or temporary variables,
e.g., \texttt{LI$\_$0}) accessed in the kernel, whereas $B'$ and $B''$
are generic problem-dependent subsets. Regardless of the specific
input problem, by abstracting from the presence of all variables
independent of both \texttt{j} and \texttt{k}, the element matrix
computation is always reducible to this form.
Figure~\ref{fig:vect-by-vect} illustrates how we can evaluate 16 entries
($j,k=0,...,3$) of the element matrix using just 2 vector registers,
which represent a 4$\times$4 tile, assuming $\vert B' \vert = \vert B''
\vert = 1$. Values in a register are shuffled each time a product is
performed. Standard compiler auto-vectorization for both GNU and Intel
compilers, instead, executes 4 broadcast operations (i.e., ``splat'' of
a value over all of the register locations) along the outer dimension to
perform the calculation. In addition to incurring a larger number of
cache accesses, it needs to keep between $f=1$ and $f=3$ extra registers
to perform the same 16 evaluations when unroll-and-jam is used, with $f$
being the unroll-and-jam factor.

@@

% FIXME: ``mesh used to discretize''
%\rule[0.2ex]{16cm}{0.5pt}
\begin{figure}[b]
\begin{center}
\includegraphics[scale=0.70]{Pictures/coffee-pipeline.pdf}
\caption{High-level view of Firedrake. COFFEE is at the core,
  receiving ASTs from a modified version of the FEniCS Form Compiler
  and producing optimized C code kernels.}
\label{fig:coffee-pipeline}
\end{center}
\end{figure}

@@

\subsubsection{Choice and Properties of the Benchmarks}
\label{sec:real-examples}
Our code transformations were evaluated in three real-world problems
based on the following PDEs: (i) Helmholtz, (ii) Advection-Diffusion, 
and (iii) Burgers.

The three chosen benchmarks are \emph{real-life kernels} and comprise
the core differential operators in some of the most frequently
encountered finite element problems in scientific computing. This is
of crucial importance because distinct problems, possibly arising  in
completely different fields, may employ (subsets of) the same differential
operators of our benchmarks, which implies similarities
and redundant patterns in the generated code. Consequently, the proposed
code transformations have a domain of applicability that goes far beyond
that of the three analyzed equations.

@@

\begin{table*}[t]
\tbl{Performance improvement due to generalized loop-invariant code
  motion over the original non-optimized code.}{ \scriptsize \licmresultsnorms }
\label{table:perf-results-licm}
\end{table*}
Table~\ref{table:perf-results-licm} illustrates the performance
improvement obtained when \emph{licm} is applied. In general, the
speedups are notable. The main reasons were anticipated in
Section~\ref{sec:code-transf-licm}: in the original code, 1)
sub-expressions invariant to outer loops are not automatically
hoisted, while 2) sub-expressions invariant to the innermost loop are
hoisted, but their execution is not auto-vectorized. These
observations come from inspection of assembly code generated by the
compiler.

@@

\begin{table*}[t]
\tbl{Performance improvement due to generalized loop-invariant code
  motion, data alignment, and padding over the original non-optimized code.}{ \scriptsize \licmapresultsnorms }
\label{table:perf-results-licmap}
\end{table*}

@@

\begin{figure}[t]
	\centering
	\subfloat[Sandy Bridge]{	\includegraphics[scale=0.48]{Pictures/helmholtz-normalized-opvect.pdf}}
	\subfloat[Xeon Phi]{	\includegraphics[scale=0.48]{Pictures/helmholtz-normalized-opvect-phi.pdf}}
	\caption{Performance improvement over \emph{licm-ap} obtained
          by \emph{op-vect} in the Helmholtz kernel. Bars suffixed
          with ``CM'' indicate that the cost model was used to
          transform the kernel.}
	\label{fig:opvect-helmholtz-speedup}
\end{figure}

\begin{figure}
	\centering
	\subfloat[Sandy Bridge]{	\includegraphics[scale=0.48]{Pictures/diffusion-normalized-opvect.pdf}}
	\subfloat[Xeon Phi]{	\includegraphics[scale=0.48]{Pictures/diffusion-normalized-opvect-phi.pdf}}
	\caption{Performance improvement over \emph{licm-ap} obtained
          by \emph{op-vect} in the Diffusion kernel. Bars suffixed
          with ``CM'' indicate that the cost model was used to
          transform the kernel.}
	\label{fig:opvect-diffusion-speedup}
\end{figure}

@@

\begin{figure}
	\centering
	\subfloat[Sandy Bridge]{	\includegraphics[scale=0.48]{Pictures/burgers-normalized-split.pdf}}
	\subfloat[Xeon Phi]{	\includegraphics[scale=0.48]{Pictures/burgers-normalized-split-phi.pdf}}
	\caption{Performance improvement over \emph{licm-ap} obtained
          by \emph{split} in the Burgers kernel. Bars suffixed with
          ``CM'' indicate that the cost model was used to transform
          the kernel.}
	\label{fig:split-burgers-speedup}
\end{figure}

@@

\begin{table}[h]
\tbl{Performance comparison between FEniCS and COFFEE on Sandy Bridge.}{
\scriptsize
\begin{tabulary}{1.0\columnwidth}{C|C|C}
\hline
Problem & Max slow down & Max speed up \\\hline
Helmholtz & - & 4.14$\times$ \\
Diffusion & - & 4.28$\times$ \\
Burgers & 2.24$\times$ & 1.61$\times$ \\\hline
\end{tabulary}
}
\label{table:comparison-to-FFC-opt}
\end{table}

@@

\begin{figure}
	\centering
	\subfloat[]{\includegraphics[scale=0.48]{Pictures/elasticity-p1.pdf}}
	\subfloat[]{\includegraphics[scale=0.48]{Pictures/elasticity-p2.pdf}}
	\caption{Performance improvement over non-optimized code for the static linear elasticity equation on a single core of the Sandy Bridge architecture.}
	\label{fig:full-app-res}
\end{figure}

@@

\section{Generality of the Approach and Applicability to Other Domains}
\label{sec:generality}
We have demonstrated that our cross-loop optimizations for arithmetic intensity are effective in the context of automated code generation for finite element local assembly. In this section, we discuss their applicability in other computational domains and, in general, their integrability within a general-purpose compiler.

COFFEE was developed as a separate, self-contained software module, with
clear input/output interfaces, rather then incorporating it within
PyOP2. This choice was motivated by two critical aspects that
characterize the generality of our research:

@@

\item The various sub-expressions in $\beta$ only depend on (i.e. iterate along) a subset of the enclosing loops. In addition, some of these sub-expressions might reduce to the same values as iterating along certain iteration spaces. This code structure motivated the generalized loop-invariant code motion technique. The intuition is that whenever sub-expressions invariant with respect to different sets of affine loops can be identified, the question of whether, where and how to hoist them, while minimizing redundant computation, arises. Pre-computation of invariant terms also increases memory requirements due to the need for temporary arrays, so it is possible that for certain architectures the transformation could actually cause slowdowns (e.g. whenever the available per-core memory is small).

@@

  transformations. For instance, knowing loop bounds allows both generation of

@@

  Section~\ref{sec:op-vect}) and discovery of redundant computation

@@

  general-purpose compiler is far from straightforward, especially if stencils are present, we believe that this could be more easily achieved in specific domains. The intuition is to map the memory access pattern onto vector registers, and then to exploit in-register shuffling to minimize the traffic between memory and processor. By demonstrating the effectiveness of multi-loop vectorization in a real scenario, our research represents an incentive for studying this technique in a broader and systematic way.
